Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts

### 2017-12-13 article/article.tex
terminado primeiro rascunho

@@

On the consideration of writing and of Toki Pona,
the considerable irrelevance of the morphsyntactic classes as
described in Section~\ref{} suggest that a appropriate coloring should
consider the syntax to identify the nouns, verbs, adjectives, adverbs,
prepositions and particles.
Also, the coloring for poems might be more appealing if considered
e.g. the counting of syllables or repeated vowels and ending
syllables.
Insights might be obtained through the observation of the
choices made and advocated in software packages such as text editors
(e.g. Vim, Emacs, Sublime, etc.),
software dedicated to syntax highlighting (e.g.
pigments \footnote{\url{http://pygments.org/docs/}} and 
highlight.js \footnote{\url{https://highlightjs.org/}},
other software (e.g. Linux terminals such as Xterm and gnome-terminal).
Finally, the careful choice of fonts is known to have
a relevant impact in comfort and productivity~\cite{fonts},
e.g. sans-sheriff fonts are more promptly read and yield a cleaner
text than a fot with sheriff.

\subsection{Wordnet Toki Pona}\label{wn}
exception: jan (people)
is a hyponym of soweli (mammal),
kili is hyponym of kasi,
(walo, pimeja, jelo, loje, laso) are hyponyms of kule.
But otherwise such relations hold:
e.g. luka (as hand) is e meronym of sijelo (body).

For a first achievement of a Toki Pona Wordnet,
each of the Toki Pona words in the dictionary
were related to Wordnet synsets
through the English lemmas.
All synsets related to the lemma with the POS tag
given in the dictionary are added.
Another immediate possibility would be to
relate each Toki Pona word to the synsets
related to all the words bounded by a semicolon.
Another possibility, which I find more consistent
but also more controversial,
is to choose all synsets related to the Wnglish
words given that the Toki Pona words can be used
in many roles.
WordNet only contains nouns, adjectives, verbs, and adverbs.
Thus, particles and prepositions were not considered.
Numbers were considered adjectives.
Words presented as adjectives in the dictionary
were considered adjectives and adverbs.

The total number of different synsets is 2493 in a very conservative
counting: discarding prepositions and particles,
only considering the synsets which have the POS tag
given in the dictionary.
If all the synsets (that have any POS tag)
related to a lemma are considered, 
then Toki Pona vocabulary reaches 3961 synsets.
The prepositions are particularly missing in these
countings because e.g. tawa or lon are used also
as ordinary words.
If synsets for them are considered those of the English
lemmas with any POS tag, the vocabulary reaches 4052
synsets.
The total number of synsets in Wordnet is X.
Total number of synsets counting neighbors.
Depth of TP synsets.
Using such features to devise other conlangs.

\subsection{Hacks from other people}
The \url{tokipona.net} has a number of tools,
just as to transliterate names into Toki Pona phonemes,
and search in corpus.

>>>>

% reference to figure of the texts.

@@

On the consideration of writing and of Toki Pona,
the considerable irrelevance of the morphsyntactic classes as
described in Section~\ref{} suggest that a appropriate coloring should
consider the syntax to identify the nouns, verbs, adjectives, adverbs,
prepositions and particles.
Also, the coloring for poems might be more appealing if considered
e.g. the counting of syllables or repeated vowels and ending
syllables.
Insights might be obtained through the observation of the
choices made and advocated in software packages such as text editors
(e.g. Vim, Emacs, Sublime, etc.),
software dedicated to syntax highlighting (e.g.
pigments \footnote{\url{http://pygments.org/docs/}} and 
highlight.js \footnote{\url{https://highlightjs.org/}},
other software (e.g. Linux terminals such as Xterm and gnome-terminal).
Finally, the careful choice of fonts is known to have
a relevant impact in comfort and productivity~\cite{fonts},
e.g. sans-sheriff fonts are more promptly read and yield a cleaner
text than a for with sheriff.

\subsection{Wordnet Toki Pona}\label{wn}
For the achievement of a first Toki Pona Wordnet,
each of the Toki Pona words in the dictionary
were related to Wordnet synsets
through the English lemmas.
All synsets related to the lemma with the POS tag
given in the dictionary are added.
Another immediate possibility would be to
relate each Toki Pona word to the synsets
simultaneously associated to all the English
words bounded by a semicolon in the dictionary.
Yet another possibility, which I find more consistent
but also more controversial,
is to choose all synsets related to the English
words given that the Toki Pona words can be used
in many roles.
Wordnet only contains nouns, adjectives, verbs, and adverbs.
Thus, particles and prepositions were not considered.
Numbers were considered adjectives.
Words presented as adjectives in the dictionary
were considered adjectives and adverbs.~\cite{wordnet}

The total number of different synsets is $2493$ in a very conservative
counting: discarding prepositions and particles,
and only considering the synsets which have the POS tag
given in the dictionary.
If all the synsets (that have any POS tag)
related to a lemma are considered, 
then Toki Pona vocabulary reaches $3961$ synsets.
The prepositions are particularly missing in such
countings because e.g. tawa or lon are used also
as ordinary words.
If synsets for them are considered those of the English
lemmas with any POS tag, the vocabulary reaches $4052$
synsets.
The total number of synsets in Wordnet is $117659$.
Total number of synsets counting neighbors is Y.
Depth of TP synsets is .
We envision the usage of such features to guide creation
of other conlangs.

The \tttt{makeWordnet.py} script, available in~\cite{tokipona},
makes available such a tentative Wordnet in the simplest form:
the Toki Pona words are keys in a dictionary that returns the
corresponding synsets.
Three of such dictionaries are implemented:
one where all synsets related to the lemma is
returned; another as such but excluding the prepositions;
one that returns only the synsets
registered in Wordnet with the same POS tag as the lemma
is in the official dictionary.

It is worth mentioning that Toki Pona words are very general
and each of them might mean many things.
Thus, the words are not very easily related e.g. by hyponymy
or meronymy.
Exceptions: jan (people) is a hyponym of soweli (mammal);
kili is an hyponym of kasi;
walo, pimeja, jelo, loje, laso
are hyponyms of kule;
luka (as hand) is a meronym of sijelo (as body).

\subsection{Hacks and resources from other people}
A number of tools and other resources for dealing with Toki Pona
were developed by the community.
The most complete list is supposedly~\cite{gdoc} and includes
videos, musical pieces, artistic texts, reference documents (e.g.
a Toki Pona - Esperanto dictionary), journalistic articles,
and software tools.
